Unsupervised Learning for Identifying Events in Active Target Experiments

### 2020-08-05 doc/Manuscript/article.tex
Updates from Overleaf

%\documentclass[preprint,12pt]{elsarticle}
% uncomment for submission to NIM A
\documentclass[review,number,sort&compress]{elsarticle}
\usepackage{lineno}
\linenumbers

@@

\begin{keyword}
Active target experiments \sep Machine Learning \sep Unsupervised Learning \sep Autoencoder Neural Networks
\PACS
%% PACS codes here, in the form: \PACS code \sep code

@@

We display two different events from the ${}^{46}$Ar experiment in figure \ref{fig:samples}. The top row illustrates a carbon event with a large fraction of noise, while the bottom row shows a proton event almost free of noise.

@@

We illustrate two filtered events in figure \ref{fig:samples_filtered}. These are the same events as shown in figure \ref{fig:samples}, but with the Hough transform and nearest neighbours filtering applied.

@@

The architecture is portrayed in Figure \ref{fig:mixae}, wherein tapered boxes denote a direction of compression in the network components. In the figure each encoder and decoder pair constitute an autoencoder. The assignment of the cluster belonging to a sample, $\mathbf{\hat{x}}$, is taken to be the index of the maximal element in the vector $\mathbf{p}$ as shown in the right-most part of the figure. Finally, the model is trained end-to-end with back propagation \cite{Linnainmaa1976}, as implemented in the machine learning library TensorFlow \cite{tensorflow}.

@@

In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in figures \ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. For example, consider the row corresponding to the proton class in Fig.~\ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

@@

To further investigate the clusters presented in the matrix in figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we study examples from the proton samples belonging to different clusters for the filtered and full data.
We display these examples in Figs.~\ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr}, respectively. Form these figures we observe that the more noisy proton events are being clustered together with the amorphous other class events. This is in line with the intuitive expectation from viewing the clustering results in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.

@@

\begin{table}[H]
\centering 
\caption[MIXAE clustering performance]{MIXAE clustering performance on the ${}^{46} Ar$ experimental data. In contrast with the VGG-16 + $K$-means approach we observe  significant variations in performance. {\color{blue}, however, the KMeans++ algorithm still clusters by reaction type, as evidenced when initializing $10$ random centroid seeds, and selecting the most separated clusters without supervision. The mean and standard deviations come from $N = 1000$ runs.  }}\label{tab:clstr_mixae}
\input{./plots/mixae_clustering.tex}
\end{table}

As with the VGG16 + $K$-Means approach we wish to further investigate the clustering results. Taking the best performing MIXAE model on filtered and raw data we tabulate the clusters against their labels. These tables are present in figures \ref{fig:mixae_confmat_filtered} and \ref{fig:mixae_confmat_raw} for filtered and raw data, respectively.

@@

In particular, the ability of the $K$-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis, where clusters of events corresponding to different reaction channels could be later identified by the experimenter.
Another interesting facet of the $K$-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. However, as can be seen from the non-proton clusters in figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw},  this does not necessarily imply that the physical signals are correspondingly clear. 
A caveat to the $K$-means method is that the number of clusters has to be specified in advance. Each experiment then has to be considered in light of possible reaction channels to determine a sensible number of clusters for this approach.

The same caveat is present in the MIXAE implementation. While it shows better optimal performance than the $K$-means method, some inconsistencies were observed, that were notably not evident from the unsupervised training-objectives of the model. These two factors currently conspire to limit its immediate applicability, and more developments are needed for this approach.

In summary, our study shows that unsupervised track classification with an implementation of the VGG16 + $K$-means approach is a viable solution. For future work, it is worth investigating whether the adaptation of  models like the MIXAE algorithm will allow better performance at no significant cost to consistency.  
%Especially considering the flexibility to use 3D input representations or auxiliary physical goals in the optimization procedure. 
The two examples of unsupervised machine learning methods studied in this work are a first encouraging step towards automated classification of events that could not only greatly reduce the resource cost of analysis, but also eventually boost the efficiency of the experiment by allowing post-trigger decisions based on such algorithm implemented in hardware. Much progress remains to be done before achieving these goals.
% End edit by DB (6/17/2020)
%From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. {\color{orange}[flip this to positive ]} It is also clear that there is significant potential in both avenues explored in this work. 
%n particular, the surprising ability of the $K$-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
%Another surprising facet of the $K$-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.

>>>>

%\documentclass[preprint,12pt]{elsarticle}
% uncomment for submission to NIM A
\documentclass[review,sort&compress]{elsarticle}
\usepackage{lineno}
%\linenumbers

@@

\begin{keyword}
Active target experiments \sep Machine Learning \sep Unsupervised Learning \sep Autoencoder Neural Networks
%\PACS
%% PACS codes here, in the form: \PACS code \sep code

@@

We display two different events from the ${}^{46}$Ar experiment in Fig.~\ref{fig:samples}. The top row illustrates a carbon event with a large fraction of noise, while the bottom row shows a proton event almost free of noise.

@@

We illustrate two filtered events in Fig.~\ref{fig:samples_filtered}. These are the same events as shown in Fig.~\ref{fig:samples}, but with the Hough transform and nearest neighbours filtering applied.

@@

The architecture is portrayed in Fig.~\ref{fig:mixae}, wherein tapered boxes denote a direction of compression in the network components. In the figure each encoder and decoder pair constitute an autoencoder. The assignment of the cluster belonging to a sample, $\mathbf{\hat{x}}$, is taken to be the index of the maximal element in the vector $\mathbf{p}$ as shown in the right-most part of the figure. Finally, the model is trained end-to-end with back propagation \cite{Linnainmaa1976}, as implemented in the machine learning library TensorFlow \cite{tensorflow}.

@@

In addition to the performance measures reported in table \ref{tab:clstr_vgg}, it is interesting to observe which samples are wrongly assigned. To investigate this problem, we compute the matrices $\mathbf{W}$ as shown in table \ref{tab:contingency}. From these tables, we can infer which classes are more or less entangled with others. The results for each data set is shown in  Figs.~\ref{fig:clster_confmat_sim} \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}. We observe that the proton class is consistently assigned in a pure cluster. For example, consider the row corresponding to the proton class in Fig.~\ref{fig:clster_confmati_filt}. The column corresponding to the largest entry in the proton row has zero other predicted classes in it.

@@

To further investigate the clusters presented in the matrix in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}, we study examples from the proton samples belonging to different clusters for the filtered and full data.
We display these examples in Figs.~\ref{fig:filtered_vgg_clster_repr} and \ref{fig:full_vgg_clster_repr}, respectively. Form these figures we observe that the more noisy proton events are being clustered together with the amorphous other class events. This is in line with the intuitive expectation from viewing the clustering results in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}.

@@

% This table, 5,  is not properly explained, what is init=10? and what is top 1? 
\begin{table}[H]
\centering 
\caption[MIXAE clustering performance]{MIXAE clustering performance on the ${}^{46} Ar$ experimental data. In contrast with the VGG-16 + $K$-means approach we observe  significant variations in performance. The $K$-means algorithm still clusters by reaction type, as evidenced when initializing $10$ random centroid seeds, and selecting the most separated clusters without supervision. The mean and standard deviations come from $N = 1000$ runs.  }\label{tab:clstr_mixae}
\input{./plots/mixae_clustering.tex}
\end{table}

As with the VGG16 + $K$-Means approach we wish to further investigate the clustering results. Taking the best performing MIXAE model on filtered and raw data we tabulate the clusters against their labels. These tables are present in Figs.~\ref{fig:mixae_confmat_filtered} and \ref{fig:mixae_confmat_raw} for filtered and raw data, respectively.

@@

In particular, the ability of the $K$-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis, where clusters of events corresponding to different reaction channels could be later identified by the experimenter.
Another interesting facet of the $K$-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. However, as can be seen from the non-proton clusters in Figs.~\ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw},  this does not necessarily imply that the physical signals are correspondingly clear. 
A caveat to the $K$-means method is that the number of clusters has to be specified in advance. Each experiment then has to be considered in light of possible reaction channels to determine a sensible number of clusters for this approach.

The same caveat is present in the MIXAE implementation. While it shows better optimal performance than the $K$-means method, some inconsistencies were observed that were notably not evident from the unsupervised training-objectives of the model. These two factors currently conspire to limit its immediate applicability, and more developments are needed for this approach.

In summary, our study shows that unsupervised track classification with an implementation of the VGG16 and the  $K$-means approach is a viable solution. For future work, it is worth investigating whether the adaptation of  models like the MIXAE algorithm will allow better performance at no significant cost to consistency.  
%Especially considering the flexibility to use 3D input representations or auxiliary physical goals in the optimization procedure. 
The two examples of unsupervised machine learning methods studied in this work are a first encouraging step towards automated classification of events that could not only greatly reduce the resource cost of analysis, but also eventually boost the efficiency of the experiment by allowing post-trigger decisions based on such algorithm implemented in hardware. Much progress remains to be done before achieving these goals.
% End edit by DB (6/17/2020)
%From our analysis it is clear that we are a ways off from a fully automated unsupervised method for track classification. {\color{orange}[flip this to positive ]} It is also clear that there is significant potential in both avenues explored in this work. 
%n particular, the surprising ability of the $K$-means algorithm in picking out clear proton clusters from the VGG16 latent space lends itself well to an exploratory phase of analysis.
%Another surprising facet of the $K$-means clustering is its consistent performance. As shown in table \ref{tab:clstr_vgg}, the variance is zero for the performance metrics. This result indicates that the clusters are very clearly defined in the VGG16 latent space. As we see from the non-proton clusters if figures \ref{fig:clster_confmati_filt} and \ref{fig:clster_confmati_raw}  that does not necessarily imply that the physical signals are correspondingly clear.
